Computing Entropies With Nested Sampling

### 2017-06-26 paper/ms.tex
A bit of writing

For experimental design purposes, if we want to maximise the expected amount
of information obtained from data, maximising either $I(\x; \d)$ or
$H(\x | \d)$ will produce the same result because the prior $p(\x)$ does not
vary with the experimental design. Reference priors
\citep{bernardo2005reference} also make use of the mutual information but in a
different way.

@@

Nested Sampling \citep[NS][]{skilling2006nested} is an algorithm whose aim is
to calculate the evidence
\begin{align}
Z &= \int \pi(\x) L(\x) \, d\x
\end{align}
where $\x$ is the unknown parameter(s), $\pi$ is the prior distribution,
and $L$ is the likelihood function. It does this evolving a population of
$N$ particles in the parameter space.
The $N$ particles are initially initialised from the prior $\pi(\x)$, and the
particle with the lowest likelihood, $L^*_1$, is found and its details are
recorded in the output. This worst particle is then discarded
and replaced by a new particle
drawn from the prior $\pi$ but subject to the constraint that its likelihood
must be above $L^*_1$. This process is repeated, resulting in an
increasing sequence of likelihoods
\begin{align}
L^*_1, L^*_2, L^*_3..., 
\end{align}

@@

The above insight, that the number of NS discarded points with likelihood
below $\ell$ has expected value $-N\ln X(\ell)$, is the basis of the algorithm.
If we wish to measure a log-probability $-\ln X(\ell)$, we can use NS to do it.
See Algorithm~\ref{alg:algorithm} for a step-by-step description of the
algorithm. The algorithm is written in terms of an unknown quantity $\x$
whose entropy is required. In applications, $\x$ may be parameters,
a subset of the parameters, data, some function of the data,
parameters and data jointly, or whatever.

@@

As the algorithm lets us calculate the entropy of any distribution which can
be sampled by MCMC, there are several options. The mutual information
can be written in the following three ways
\begin{align}
I(x; y) &= H(x) + H(y) - H(x, y) \\
I(x; y) &= H(y) - H(y|x) \\
I(x; y) &= H(x) - H(x|y).
\end{align}
In a Bayesian problem where $x$ is an unknown parameter and $y$ is data, the
first and second formulas would be very costly to compute, because of the
high dimensional probability distribution for the dataset.

>>>>

For experimental design purposes, if we want to maximise the expected amount
of information obtained from data, maximising either $I(\x; \d)$ or
$H(\x | \d)$ will produce the same result because the prior $p(\x)$ does not
vary with the experimental design. Reference priors
\citep{bernardo2005reference} also maximise $I(\x; \d)$ but vary the prior
in the maximisation process.

@@

Nested Sampling \citep[NS][]{skilling2006nested} is an algorithm whose aim is
to calculate the evidence
\begin{align}
Z &= \int \pi(\x) L(\x) \, d\x
\end{align}
where $\x$ is the unknown parameter(s), $\pi$ is the prior distribution,
and $L$ is the likelihood function. While $Z$ is a simple expectation, the
implied distribution of $L$-values tends to be very heavy-tailed, which is
why simple Monte Carlo does not work. Equivalently, the integral is dominated
by very small regions where $L(\x)$ is high.

NS evolves a population of
$N$ particles in the parameter space.
The particles are initially initialised from the prior $\pi(\x)$, and the
particle with the lowest likelihood, $L^*_1$, is found and its details are
recorded as output. This worst particle is then discarded
and replaced by a new particle
drawn from the prior $\pi$ but subject to the constraint that its likelihood
must be above $L^*_1$ (this is usually achieved by cloning a surviving particle
and evolving it with MCMC). This process is repeated, resulting in an
increasing sequence of likelihoods
\begin{align}
L^*_1, L^*_2, L^*_3..., 
\end{align}

@@

The above insight, that the number of NS discarded points with likelihood
below $\ell$ has expected value $-N\ln X(\ell)$, is the basis of the algorithm.
If we wish to measure a log-probability $-\ln X(\ell)$, we can use NS to do it.
$\pi(\x)$ and $L(\x)$ need not be the prior and likelihood respectively, but
can be any probability distribution (over any space) and any function whose
expected value is needed for a particular application.

See Algorithm~\ref{alg:algorithm} for a step-by-step description of the
algorithm. The algorithm is written in terms of an unknown quantity $\x$
whose entropy is required. In applications, $\x$ may be parameters,
a subset of the parameters, data, some function of the data,
parameters and data jointly, or whatever.

@@

As the algorithm lets us calculate the entropy of any distribution which can
be sampled by MCMC, there are several options. The mutual information
can be written in the following three ways
\begin{align}
I(x; y) &= H(x) + H(y) - H(x, y) \\
I(x; y) &= H(y) - H(y|x) \\
I(x; y) &= H(x) - H(x|y) \\
I(x; y) &= \int D_{\rm KL}\left(p(x|y) || p(x)\right) \, dy.
\end{align}
In a Bayesian problem where $x$ is an unknown parameter and $y$ is data, the
first and second formulas would be very costly to compute, because of the
high dimensional probability distribution for the dataset.
